### 2015-04-18 conclusion.tex
Submitted version

\todo{.5 pg}

>>>>

The complexity of managing HPC software is rapidly increasing, and it
will continue unabated without effective tools.
In this paper, we reviewed the state of software management tools 
across a number of HPC sites, with particular focus on Livermore
Computing (LC). While tools exist that can handle multi-configuration
installs, none of them addresses the combinatorial nature
of the software configuration space directly. None of them allows
a user to rapidly {\it compose} new parametric builds.

We introduced Spack, a package manager in development at LLNL, that
provides truly {\it parameterized} builds.  Spack implements
a novel, recursive {\it spec} syntax that simplifies the process of working
with large software configuration spaces, and it builds software 
so that it will run correctly, regardless of the environment.
We outlined a number of additional unique features of Spack, including
versioned virtual dependencies, and Spack's {\it concretization} process,
which converts an underspecified build DAG into a buildable spec.
We showed through several use cases that Spack is already
increasing operational efficiency and simplifying software management
at LLNL, and that it can be rapidly adapted to new production use cases.

### 2015-04-18 implementation.tex
Added versioned vdeps
GLL edits
Updated concretization section.
Submitted version

\section{The Spack Package Manager}
\label{sec:implementation}
Based on our experiences at LLNL, we have developed
{\it Spack}, the Supercomputing PACKage manager.
Spack is written in Python.  We chose Python for its flexibility
its wide use in the HPC community.
%
Like prior systems, Spack supports an arbitrary number of software
installations, and like Nix it can identify them with hashes.  Unlike any
prior system, Spack provides a language to specify and manage the
combinatorial space of HPC software configurations.

\noindent
Spack has the following unique capabilities:
\begin{enumerate}
\item Packages are {\bf parametrized}, allowing new builds to be composed
      on the fly for different versions, architecture, compiler, options, 
      and dependencies.
\item To manage this large parameter space, spack provides a novel, 
      recursive {\bf spec syntax} for dependency graphs.
\item Specs support versioned, ABI-incompatible interfaces like MPI through
      {\bf versioned virtual dependencies}.
\item Spack builds with {\bf compiler wrappers} that enforce build
      consistency and simplify package writing.
\end{enumerate}

\input{implementation-packages}
\input{implementation-specs}
\input{implementation-virtual-deps}
\input{implementation-concretization}

\input{implementation-environment}

>>>>

\section{The Spack Package Manager}
\label{sec:implementation}
Based on our experiences at LLNL, we have developed
{\it Spack}, the Supercomputing PACKage manager.
Spack is written in Python.  We chose Python for its flexibility
and its wide use in the HPC community.
%
Like prior systems, Spack supports an arbitrary number of software
installations, and like Nix it can identify them with hashes.  Unlike any
prior system, Spack provides a language to specify and manage the
combinatorial space of HPC software configurations.

\noindent
Spack has the following unique capabilities:
\begin{enumerate}
\item Packages are {\bf parameterized}, allowing new builds to be composed
      on the fly for different versions, architecture, compiler, options, 
      and dependencies.
\item To manage this large parameter space, spack provides a novel, 
      recursive {\bf spec syntax} for dependency graphs.
\item Specs support versioned, ABI-incompatible interfaces like MPI through
      {\bf versioned virtual dependencies}.
\item Spack builds with {\bf compiler wrappers} that enforce build
      consistency and simplify package writing.
\end{enumerate}

\input{implementation-packages}
\input{implementation-specs}
\input{implementation-virtual-deps}
\input{implementation-concretization}
\input{implementation-environment}

### 2015-04-18 motivation.tex
Trim a bit.
neg vspace in table 1 and 2
Submitted version

To cope with software diversity, many HPC sites use a combination of existing
package managers and either manual or semi-automated software installation.  
For the baseline OS, many sites maintain traditional binary
packages using the vendor's package manager. LLNL maintains a Linux 
distribution, CHAOS~\cite{chaos} for this purpose, which is managed by RPM.

For more specific builds, many sites adhere to detailed naming conventions
that encode information in file system paths. 
Table~\ref{tab:naming-conventions} shows a sampling of conventions
at several sites. At LLNL, we use the APT package manager for installs 
in the {\tt /usr/local/tools} file system; we use {\tt /usr/global/tools} 
for manual installs.
ORNL uses hand installs but adheres to a strict scripting structure
that can be used to reproduce each build~\cite{jones+:cug08}.
TACC relies heavily on locally maintained RPMs for its software installs.

@@

\paragraph{Environment Modules and RPATHs}

>>>>

To cope with software diversity, many HPC sites use a combination of existing
package managers and either manual or semi-automated installation.
For the baseline OS, many sites maintain traditional binary
packages using the vendor's package manager. LLNL maintains a Linux
distribution, CHAOS~\cite{chaos} for this purpose, which is managed by RPM.

For more specific builds, many sites adhere to detailed naming conventions
that encode information in file system paths.
Table~\ref{tab:naming-conventions} shows conventions
at several sites. At LLNL, we use the APT package manager for installs
in the {\tt /usr/local/tools} file system; we use {\tt /usr/global/tools}
for manual installs.
ORNL uses hand installs but adheres to a strict scripting structure
to reproduce each build~\cite{jones+:cug08}.
TACC relies heavily on locally maintained RPMs for its installs.

@@

\paragraph{Environment Modules and RPATHs}\label{sec:env-rpath}

### 2015-04-18 usecase-policy.tex
Submitted version

While Spack can easily create a dozen installations of a package like {\tt mpileaks}
for different compilers and MPI implementations,
it can be difficult for an end-user to determine which of those dozen packages should be used with their application.
As discussed in Section~\ref{sec:envmodule},
environment modules are commonly used on supercomputer systems to solve this problem,
and Spack allows the package administrator to automatically create module and dotkit files for packages that Spack installs.

However, even when modules and dotkits are available,
many users will go directly through the filesystem to access package installs.
Spack installs packages into paths based on concretized specs,
which is ideal for maintaining multiple package installations, but may be difficult for an end-user to navigate.
A version of {\tt mpileaks}, for example, may be installed in a location like
%
\begin{minted}[fontsize=\small]{bash}
    spack/opt/chaos_5_x86_64_ib/ gcc@4.9.2/mpileaks@1.0-db465029
\end{minted}
%
Spack thus allows the creation of views, which are a symbolic-link based directory layout of packages.
Views provide a human-readable directory layout,
which can be taylored to fit alongside existing directory layouts or used to create new layouts.
For example, the above {\tt mpileaks} package may have a view that creates a symlink in
{\tt /opt/mpileaks-1.0-openmpi} to the Spack installation of {\tt mpilleaks}.
The same package install may be referenced by multiple symlinks and views,
so the above package could also be symlinked from a more generic {\tt /opt/mpileaks-openmpi}
symlink (which is useful for users who don't want to hardcode a specific version).
Views can also be used to create symlinks to specific executables or libraries in an install,
so a Spack-built {\tt gcc 4.9.2} install may have a view that creates symlinks from
{\tt /bin/gcc49} and {\tt /bin/g++49} to the appropriate {\tt gcc} and {\tt g++} executables.

Views are configured through a {\tt spackconfig} file, which can be setup at a site-wide level or for individual users.
For each package or set of packages, the {\tt spackconfig} file contains rules
describing the symlinks that should point into that package.
The symlink names can be parameterized.
For example, the above {\tt mpileaks} symlink might have been created by rule like
%
\begin{minted}[fontsize=\small]{bash}
    /opt/${PACKAGE}-${VERSION}-${MPINAME}
\end{minted}
%
When Spack installs or removes packages,
symlinks are automaticaly created, deleted, or updated according to these rules.

Spack's views are a projection from a point in a high-dimension space
(the concretized spec, which fully specifies all parameters) to point in a lower-dimension space
(the symlink name, which may only specify a few parameters).
Thus several package installations may map to the same symlink.
For example, the above {\tt mpileaks} symlink could point to an {\tt mpileaks} compiled with
{\tt gcc} or {\tt icc} -- the compiler parameter is not part of the symlink.
To keep  package installations  consistent and reproducible,
Spack has a well-defined mechanism for resolving conflicting symlinks.
Spack defines a sort order for packages, which arranges packages from ``most desired'' to ``least desired''.
When a symlink could resolve to multiple package installations,
Spack points the symlink at the most desired package.

By default, Spack treats newer versions of packages compiled with newer compilers
as more desirable than older packages built with older compilers.
It has well-defined, but not necessarily meaningful,
preferences for deciding between MPI implementations and different compilers.
A user can override Spack's default sort order through their {\tt spackconfig} file.
For example, at one site users may typically use the Intel compiler,
but some users also use the system's default {\tt gcc 4.4.7}.
These preferences could be stated by adding:
%
\begin{minted}[fontsize=\small]{bash}
    compiler_order = icc,gcc@4.4.7
\end{minted}
%
to the site's {\tt spackconfig} file, which would cause the ambigious {\tt mpileaks} symlink to point to an {\tt icc} installation.
Any compiler left out of the compiler\_order setting is treated as less preferred than explicitly stated compilers.
Spack can also be configured to treat specific package versions as more preferred than other versions,
which can be useful if a new version is unstable and untested, and should not be advertised with symlinks.

@@

To support our own private packages, and to support those of LLNL code teams, Spack allows the creation of site-specific variants of packages.  Via the {\tt spackconfig} file, users can specify additional search directories for finding additional {\tt Package} classes.
%
The additional packages are like the {\tt mpileaks} package shown in Figure~\ref{fig:mpileaks}.  However, the extension packages can extend from not only {\tt Package}, but also any of Spack's built-in packages.   Custom packages can inherit from and replace the Spack's default packages, allowing sites to either tweak or completely replace Spack's build recipes.  To continue the previous example, a site can write a {\tt LocalSpindle} Python class, which inherits from Spack's {\tt Spindle} class.  {\tt LocalSpindle} may simply add additional configure flags to the {\tt Spindle} class, while leaving the dependencies and most of the build instructions from its parent class.  To keep builds deterministic and reproducable, Spack also tracks which {\tt Package} class drove a specific build.

>>>>

While Spack can easily create many installations of a package like {\tt mpileaks}
for different compilers and MPI implementations, end users may find Spack's directory layout
confusing, and they may not be able to find the right libraries.
As discussed in Section~\ref{sec:envmodule},
environment modules are commonly used on supercomputer systems to solve this problem,
and Spack allows the package author to automatically create module and dotkit files for packages that Spack installs.

However, even when modules and dotkits are available,
many users will go directly through the filesystem to access package installs.
Spack installs packages into paths based on concretized specs,
which is ideal for maintaining multiple package installations, but may be difficult for an end-user to navigate.
A version of {\tt mpileaks}, for example, may be installed in a location like
%
\begin{minted}[fontsize=\small]{bash}
spack/opt/chaos_5_x86_64_ib/gcc@4.9.2/mpileaks@1.0-db465029
\end{minted}
%
Spack thus allows the creation of views, which are a symbolic-link based directory layout of packages.
Views provide a human-readable directory layout,
which can be taylored to fit alongside existing directory layouts or used to create new layouts.
For example, the above {\tt mpileaks} package may have a view that creates a link in
{\tt /opt/mpileaks-1.0-openmpi} to the Spack installation of {\tt mpilleaks}.
The same package install may be referenced by multiple links and views,
so the above package could also be linked from a more generic {\tt /opt/mpileaks-openmpi}
link (which is useful for users who don't want to hardcode a specific version).
Views can also be used to create symbolic links to specific executables or libraries in an install,
so a Spack-built {\tt gcc 4.9.2} install may have a view that creates links from
{\tt /bin/gcc49} and {\tt /bin/g++49} to the appropriate {\tt gcc} and {\tt g++} executables.

Views are configured through a {\tt spackconfig} file, which can be set up at the site or user level.
For each package or set of packages, the {\tt spackconfig} file contains rules
describing the links that should point into that package.
The link names can be parameterized.
For example, the above {\tt mpileaks} symbolic link might have been created by a rule like:
%
\begin{minted}[fontsize=\small]{bash}
    /opt/${PACKAGE}-${VERSION}-${MPINAME}
\end{minted}
%
On installation or removal,
links are automatically created, deleted, or updated according to these rules.

Spack's views are a projection from a point in a high-dimension space
(the concretized spec, which fully specifies all parameters) to point in a lower-dimension space
(the link name, which may only specify a few parameters).
Thus several package installations may map to the same link.
For example, the above {\tt mpileaks} link could point to an {\tt mpileaks} compiled with
{\tt gcc} or {\tt icc} -- the compiler parameter is not part of the link.
To keep  package installations  consistent and reproducible,
Spack has a well-defined mechanism for resolving conflicting links.
Spack defines a sort order for packages, which arranges packages from ``most desired'' to ``least desired''.
When a symbolic link could resolve to multiple package installations,
Spack points the link at the most desired package.

By default, Spack treats newer versions of packages compiled with newer compilers
as more desirable than older packages built with older compilers.
It has well-defined, but not necessarily meaningful,
preferences for deciding between MPI implementations and different compilers.
A user can override Spack's default sort order through their {\tt spackconfig} file.
For example, at one site users may typically use the Intel compiler,
but some users also use the system's default {\tt gcc 4.4.7}.
These preferences could be stated by adding:
%
\begin{minted}[fontsize=\small]{bash}
    compiler_order = icc,gcc@4.4.7
\end{minted}
%
to the site's {\tt spackconfig} file, which would cause the ambigious {\tt mpileaks} link to point to an {\tt icc} installation.
Any compiler left out of the compiler\_order setting is treated as less preferred than explicitly stated compilers.
Spack can also be configured to treat specific package versions as more preferred than other versions,
which can be useful if a new version is unstable and untested, and should not be advertised with symbolic links.

@@

To support our own private packages, and to support those of LLNL code teams, Spack allows the creation of site-specific variants of packages.  Via the {\tt spackconfig} file, users can specify additional search directories for finding additional {\tt Package} classes.
%
The additional packages are like the {\tt mpileaks} package shown in Figure~\ref{fig:mpileaks}.  However, the extension packages can extend from not only {\tt Package}, but also any of Spack's built-in packages.   Custom packages can inherit from and replace the Spack's default packages, allowing sites to either tweak or completely replace Spack's build recipes.  To continue the previous example, a site can write a {\tt LocalSpindle} Python class, which inherits from Spack's {\tt Spindle} class.  {\tt LocalSpindle} may simply add additional configure flags to the {\tt Spindle} class, while leaving the dependencies and most of the build instructions from its parent class.  For reproducibility, Spack also tracks the {\tt Package} class that drove a specific build.

### 2015-04-18 usecase-python.tex
Submitted version

The end result is that Python extensions can now be installed automatically
in their own prefixes, but they can be composed with a wide range
of bleeding-edge libraries that other package managers do not handle.  To experiment
with these extensions, users can load environment modules generated by Spack.
If they want a particular version to be available without any special environment 
settings, they can activate it within the Python instance.

>>>>

The end result is that Python extensions can now be installed automatically
in their own prefixes, and they can be composed with a wide range
of bleeding-edge libraries that other package managers do not handle.  To experiment
with these extensions, users can load environment modules generated by Spack.
If they want a particular version to be available without any special environment 
settings, they can activate it within the Python instance.
